---
title: "Insulin Pharmacokinetics Model: Injection Site Effects"
subtitle: "Three-Compartment PBPK Simulation — EEB 119 Project"
author: "Aditya Narashim, Richemond Shin, Nancy Robles-Perez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
# ============================================================================
# SETUP CHUNK
# ----------------------------------------------------------------------------
# This chunk loads all required packages and sets global options for the
# document. All plots will use a consistent theme and sizing.
# ============================================================================

knitr::opts_chunk$set(
  echo    = TRUE,     # Show code in output
  warning = FALSE,    # Suppress warnings
  message = FALSE,    # Suppress package loading messages
  fig.align = "center"
)

# --- Required packages ---
# deSolve: numerical ODE integration (lsoda solver, handles stiff systems)
# ggplot2: grammar-of-graphics plotting
# dplyr/tidyr: data wrangling
# patchwork: composing multi-panel figures
# plotly: interactive 3D phase portrait
# knitr/kableExtra: formatted tables
install.packages("deSolve")
library(deSolve)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(plotly)
library(knitr)
library(kableExtra)
```

# 1. Model Overview

This document implements the three-compartment physiologically-based pharmacokinetic
(PBPK) model described in our project proposal. The model tracks:
  
- **S(t)**: Insulin concentration in the subcutaneous depot (injection site)
- **P(t)**: Active insulin concentration in blood plasma
- **G(t)**: Blood glucose concentration

The central question is: *how does injection site physiology (abdomen, arm, thigh, 
buttock) affect the pharmacokinetic profile of insulin — specifically peak plasma 
concentration (C~max~), time-to-peak (T~max~), and total systemic exposure (AUC)?*

Injection site enters the model through the absorption function parameters (V~max~, K~m~),
which vary by site due to differences in local blood flow, tissue thickness, and 
capillary density.

---

# 2. Model Equations

The system of ordinary differential equations (ODEs):

$$\frac{dS}{dt} = u(t) - \frac{V_{max} \cdot S}{K_m + S} - k_{deg} \cdot S + k_{re} \cdot P$$

$$\frac{dP}{dt} = \frac{1}{V_c}\left[\frac{V_{max} \cdot S}{K_m + S} + E(G) \cdot C(t)\right] - (k_{clr} + k_{enz} + k_{re}) \cdot P$$

$$\frac{dG}{dt} = G_b + G_{in}(t) - \frac{U_{max} \cdot P \cdot G}{(K_P + P)(K_G + G)}$$

where the auxiliary functions are:

$$E(G) = \frac{E_{max} \cdot G^n}{EC_{50}^n + G^n} \quad \text{(Hill-type β-cell secretion)}$$

$$C(t) = \frac{1}{1 + P / IC_{50}} \quad \text{(negative feedback inhibition)}$$

---

# 3. Parameter Definitions

```{r parameters}
# ============================================================================
# PARAMETER DEFINITIONS
# ============================================================================
# All parameters are collected into a named list. Each is annotated with its
# physiological meaning, units, and source rationale.
#
# Parameters are organized into functional groups:
#   (A) Absorption — site-dependent, controls insulin movement from depot to plasma
#   (B) Degradation & clearance — controls insulin removal from each compartment
#   (C) Distribution — plasma volume
#   (D) Endogenous insulin — β-cell secretion response to glucose
#   (E) Glucose dynamics — basal production, insulin-mediated uptake
# ============================================================================

base_params <- list(
  
  # --- (A) ABSORPTION PARAMETERS (site-dependent) ---
  # These are the key parameters that VARY by injection site.
  # Vmax: maximum absorption rate from SC depot into plasma (Units/hr)
  #        Higher Vmax = faster absorption = higher blood flow at site
  # Km:   Michaelis-Menten half-saturation constant (Units)
  #        Lower Km = absorption saturates at lower depot concentrations
  #        Together, kabs(S) = Vmax * S / (Km + S) is a saturating function:
  #          - At low S: kabs ≈ (Vmax/Km) * S  (first-order, linear)
  #          - At high S: kabs ≈ Vmax           (zero-order, saturated)
  Vmax = 18,    # Will be overridden per site
  Km   = 30,    # Will be overridden per site
  
  # --- (B) DEGRADATION & CLEARANCE PARAMETERS ---
  # kdeg: first-order degradation rate of insulin in the SC depot (1/hr)
  #       Represents local enzymatic breakdown before absorption into blood.
  #       Insulin that degrades in the depot never reaches the plasma.
  kdeg = 0.05,
  
  # kre: redistribution/reabsorption rate constant (1/hr)
  #      Represents insulin moving FROM plasma BACK to SC tissue.
  #      Small but included for mass-balance completeness.
  kre = 0.01,
  
  # kclr: renal + hepatic clearance rate of insulin from plasma (1/hr)
  #       Primary route of insulin elimination. The liver clears ~60% of
  #       portal insulin; kidneys clear ~35% of peripheral insulin.
  kclr = 0.15,
  
  # kenz: enzymatic degradation rate in plasma (1/hr)
  #       Insulin-degrading enzyme (IDE) in blood and peripheral tissues.
  #       Separate from hepatic/renal clearance.
  kenz = 0.08,
  
  # --- (C) DISTRIBUTION ---
  # Vc: apparent volume of distribution for insulin in plasma (L)
  #     Converts mass flux (from absorption) into concentration.
  #     Typical value ~5 L for a 70 kg adult (insulin distributes mainly
  #     in extracellular fluid).
  Vc = 5.0,
  
  # --- (D) ENDOGENOUS INSULIN SECRETION ---
  # These parameters define E(G), the Hill-type function describing how
  # pancreatic β-cells secrete insulin in response to blood glucose.
  #
  # Emax: maximum endogenous insulin secretion rate (Units/hr)
  #       Upper bound of β-cell output when glucose is very high.
  Emax = 3.0,
  
  # EC50: glucose concentration at half-maximal secretion (mg/dL)
  #       The glucose level at which β-cells produce 50% of Emax.
  EC50 = 120,
  
  # hillN: Hill coefficient (dimensionless)
  #        Controls the steepness/switch-like behavior of the secretion
  #        response. Higher n = sharper threshold. n=2.5 gives a
  #        moderately sigmoidal response curve.
  hillN = 2.5,
  
  # IC50: insulin concentration for 50% inhibition of secretion (Units/L)
  #       Part of the negative feedback: C(t) = 1/(1 + P/IC50).
  #       When plasma insulin P = IC50, endogenous secretion is halved.
  #       This prevents runaway positive feedback in the P ↔ G loop.
  IC50 = 25,
  
  # --- (E) GLUCOSE DYNAMICS ---
  # Gb: basal (fasting) hepatic glucose production rate (mg/dL/hr)
  #     The liver continuously produces glucose via gluconeogenesis and
  #     glycogenolysis. This is the "always on" glucose source.
  Gb = 1.8,
  
  # Umax: maximum insulin-mediated glucose utilization rate (mg/dL/hr)
  #       Upper bound of glucose uptake by peripheral tissues (muscle, fat).
  Umax = 5.5,
  
  # KP: half-saturation constant for insulin in utilization (Units/L)
  #     The plasma insulin level at which utilization is 50% of max
  #     (holding glucose constant). Controls insulin sensitivity.
  KP = 15,
  
  # KG: half-saturation constant for glucose in utilization (mg/dL)
  #     The glucose level at which utilization is 50% of max
  #     (holding insulin constant). Controls glucose sensitivity.
  KG = 80
)
```

```{r site_profiles}
# ============================================================================
# INJECTION SITE PROFILES
# ============================================================================
# Each injection site has a different local physiology that affects how quickly
# insulin is absorbed from the subcutaneous depot into the bloodstream.
#
# Key physiological differences:
#   - Abdomen: highest capillary density and blood flow → fastest absorption
#   - Upper arm: intermediate blood flow
#   - Buttock: moderate blood flow, thicker SC layer
#   - Thigh: lowest blood flow of common sites → slowest absorption
#
# These differences are captured by varying Vmax and Km:
#   - Higher Vmax = faster maximum absorption (more blood flow)
#   - Lower Km = absorption activates at lower depot levels
#
# Reference: Gradel et al. (2018), Koivisto & Felig (1980)
# ============================================================================

site_profiles <- data.frame(
  site  = c("Abdomen", "Upper Arm", "Buttock", "Thigh"),
  Vmax  = c(18, 13, 10, 8),
  Km    = c(30, 40, 42, 50),
  color = c("#c53030", "#d69e2e", "#2b6cb0", "#718096"),
  stringsAsFactors = FALSE
)

site_profiles %>%
  kable(
    caption = "Injection Site Absorption Parameters",
    col.names = c("Site", "V_max (U/hr)", "K_m (U)", "Plot Color")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 4. ODE System Implementation

```{r ode_system}
# ============================================================================
# ODE SYSTEM FUNCTION
# ============================================================================
# This function defines the right-hand side of the ODE system.
# It is called repeatedly by the deSolve integrator at each time step.
#
# Arguments:
#   t      - current time (hours)
#   state  - named vector of current state variable values: c(S, P, G)
#   params - named list of all model parameters
#
# Returns:
#   A list containing a vector of derivatives: c(dS/dt, dP/dt, dG/dt)
#   and a vector of auxiliary/diagnostic variables for post-hoc analysis.
# ============================================================================

insulin_ode <- function(t, state, params) {
  with(as.list(c(state, params)), {
    
    # ------------------------------------------------------------------
    # EXOGENOUS INPUTS
    # ------------------------------------------------------------------
    
    # Injection input u(t):
    # Each dose is modeled as a brief square pulse over a 3-minute window
    # (0.05 hr). This avoids a true Dirac delta (which is numerically
    # problematic) while still approximating a rapid bolus injection.
    # The pulse height = dose_amount / injection_duration so that the
    # integral over the pulse equals the total dose.
    u_input <- 0
    for (i in seq_len(nrow(dose_schedule))) {
      t_dose    <- dose_schedule$time[i]
      dose_amt  <- dose_schedule$amount[i]
      inj_dur   <- 0.05  # 3-minute injection window (hours)
      if (t >= t_dose && t < t_dose + inj_dur) {
        u_input <- u_input + dose_amt / inj_dur
      }
    }
    
    # Meal glucose input Gin(t):
    # Each meal is modeled as a gamma-shaped absorption profile:
    #   Gin(t) = M * (t_rel / t_peak) * exp(-(t_rel - t_peak) / t_decay)
    # where t_rel is time since meal start, t_peak is time of peak
    # absorption (~18 min = 30% of meal duration), and t_decay controls
    # the tail. This captures the physiological pattern: glucose rises
    # after eating, peaks ~20-30 min post-meal, then tapers off.
    G_in <- 0
    for (i in seq_len(nrow(meal_schedule))) {
      t_meal   <- meal_schedule$time[i]
      meal_amt <- meal_schedule$amount[i]
      meal_dur <- meal_schedule$duration[i]
      t_rel    <- t - t_meal
      if (t_rel >= 0 && t_rel < meal_dur) {
        t_peak <- meal_dur * 0.3
        t_decay <- meal_dur * 0.4
        G_in <- G_in + meal_amt * (t_rel / t_peak) * exp(-(t_rel - t_peak) / t_decay)
      }
    }
    
    # ------------------------------------------------------------------
    # AUXILIARY FUNCTIONS
    # ------------------------------------------------------------------
    
    # Ensure non-negative state values (numerical safety)
    S_pos <- max(S, 0)
    P_pos <- max(P, 0)
    G_pos <- max(G, 0)
    
    # Saturating absorption (Michaelis-Menten kinetics):
    # kabs(S) = Vmax * S / (Km + S)
    # At low depot levels: approximately first-order (linear in S)
    # At high depot levels: saturates at Vmax (zero-order)
    # This nonlinearity is important because immediately after injection,
    # the depot is loaded (high S → saturated absorption), then as insulin
    # drains, absorption slows nonlinearly.
    kabs <- Vmax * S_pos / (Km + S_pos)
    
    # Endogenous insulin secretion E(G) — Hill function:
    # E(G) = Emax * G^n / (EC50^n + G^n)
    # This is a sigmoidal function of glucose. β-cells secrete more insulin
    # when blood glucose is high. The Hill coefficient (n=2.5) makes this
    # a moderately switch-like response — there's a threshold glucose level
    # below which secretion is minimal and above which it ramps up steeply.
    G_pow <- G_pos^hillN
    E_G   <- Emax * G_pow / (EC50^hillN + G_pow)
    
    # Inhibitory feedback factor C(t):
    # C(t) = 1 / (1 + P / IC50)
    # When plasma insulin is low:  C ≈ 1 (no inhibition, full secretion)
    # When plasma insulin is high: C → 0 (secretion suppressed)
    # This prevents a dangerous positive feedback loop:
    #   high G → high E(G) → high P → low G → low E(G) → etc.
    # Without C(t), the system could oscillate unrealistically.
    C_t <- 1 / (1 + P_pos / IC50)
    
    # Insulin-mediated glucose utilization U(P, G):
    # U(P,G) = Umax * P * G / ((KP + P)(KG + G))
    # This is a dual-substrate Michaelis-Menten form. Glucose uptake by
    # tissues (muscle, fat, liver) requires BOTH insulin AND glucose to be
    # present. It saturates independently in each:
    #   - At very high P: U ≈ Umax * G / (KG + G)  (insulin-saturated)
    #   - At very high G: U ≈ Umax * P / (KP + P)  (glucose-saturated)
    # This coupling is the fundamental insulin-glucose interaction.
    U_PG <- Umax * P_pos * G_pos / ((KP + P_pos) * (KG + G_pos))
    
    # ------------------------------------------------------------------
    # DIFFERENTIAL EQUATIONS
    # ------------------------------------------------------------------
    
    # dS/dt: Rate of change of insulin in the subcutaneous depot
    #   (+) u_input:   exogenous insulin arriving from injection
    #   (-) kabs:      insulin absorbed from depot into plasma
    #   (-) kdeg * S:  insulin degraded locally in the depot (never reaches blood)
    #   (+) kre * P:   small redistribution of insulin from plasma back to depot
    dS <- u_input - kabs - kdeg * S_pos + kre * P_pos
    
    # dP/dt: Rate of change of active insulin in plasma
    #   (+) kabs / Vc:         absorbed insulin entering plasma (divided by volume)
    #   (+) E_G * C_t / Vc:    endogenous β-cell secretion (feedback-modulated)
    #   (-) kclr * P:          renal + hepatic clearance
    #   (-) kenz * P:          enzymatic degradation (IDE)
    #   (-) kre * P:           redistribution back to SC tissue
    dP <- kabs / Vc + E_G * C_t / Vc - kclr * P_pos - kenz * P_pos - kre * P_pos
    
    # dG/dt: Rate of change of blood glucose
    #   (+) Gb:    basal hepatic glucose production (constant)
    #   (+) G_in:  glucose absorbed from food (time-varying)
    #   (-) U_PG:  insulin-mediated glucose uptake by tissues
    dG <- Gb + G_in - U_PG
    
    # ------------------------------------------------------------------
    # RETURN
    # ------------------------------------------------------------------
    # Return derivatives AND auxiliary variables for diagnostics.
    # The auxiliary variables are not part of the ODE state but are useful
    # for plotting absorption rates, secretion rates, etc.
    return(list(
      c(dS, dP, dG),                            # derivatives
      kabs   = kabs,                             # absorption rate
      E_G    = E_G,                              # endogenous secretion rate
      C_t    = C_t,                              # inhibitory factor
      U_PG   = U_PG,                             # glucose utilization rate
      G_in   = G_in,                             # meal glucose input
      u_input = u_input                          # injection input
    ))
  })
}
```

---

# 5. Dosing & Meal Schedules

```{r schedules}
# ============================================================================
# DOSING SCHEDULE
# ============================================================================
# Defines when and how much insulin is injected. Each row is one injection.
#
# Columns:
#   time   - injection time (hours from simulation start)
#   amount - dose size (Units of insulin)
#
# This schedule represents a typical 3-meal bolus regimen:
#   - Pre-breakfast bolus at t=0.5 hr
#   - Pre-lunch bolus at t=6 hr
#   - Pre-dinner bolus at t=12 hr
#   - Optional evening correction at t=18 hr
# ============================================================================

dose_schedule <- data.frame(
  time   = c(0.5, 6.0, 12.0, 18.0),
  amount = c(10,  8,   10,   5)
)

# ============================================================================
# MEAL SCHEDULE
# ============================================================================
# Defines when and how much carbohydrate (glucose equivalent) is consumed.
#
# Columns:
#   time     - meal start time (hours)
#   amount   - carbohydrate load, scaled to glucose appearance rate (mg/dL/hr)
#   duration - meal absorption window (hours); glucose appears over this period
#              with a gamma-shaped profile peaking at ~30% of the duration.
#
# Meals are timed ~15 min after each insulin dose to reflect the typical
# recommendation to inject before eating.
# ============================================================================

meal_schedule <- data.frame(
  time     = c(0.75, 6.25, 12.25, 18.5),
  amount   = c(50,   40,   55,    20),
  duration = c(0.75, 0.75, 1.0,   0.5)
)
```

```{r display_schedules}
# Display schedules as formatted tables
dose_schedule %>%
  kable(caption = "Insulin Dosing Schedule",
        col.names = c("Time (hr)", "Dose (Units)")) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)

meal_schedule %>%
  kable(caption = "Meal (Glucose Intake) Schedule",
        col.names = c("Time (hr)", "Carb Load", "Duration (hr)")) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

---

# 6. Run Simulations Across All Injection Sites

```{r run_simulations}
# ============================================================================
# SIMULATION LOOP
# ============================================================================
# We run the ODE solver once for each injection site. The ONLY parameters that
# change between runs are Vmax and Km (the absorption parameters). Everything
# else — clearance, secretion, glucose dynamics, dosing, meals — is held
# constant. This isolates the injection site effect.
#
# Solver: lsoda (from deSolve) — an adaptive-step solver that automatically
# handles stiff systems. It switches between Adams (non-stiff) and BDF (stiff)
# methods as needed.
#
# Initial conditions:
#   S(0) = 0     : empty depot (no insulin injected yet)
#   P(0) = 0     : no exogenous insulin in plasma yet
#   G(0) = 90    : fasting blood glucose ~90 mg/dL (normal range: 70-100)
# ============================================================================

# Time vector: simulate 24 hours at fine resolution
times <- seq(0, 24, by = 0.01)  # 0.01 hr = 36 sec resolution

# Initial state
state_init <- c(S = 0, P = 0, G = 90)

# Storage for all results
all_results <- list()

for (i in 1:nrow(site_profiles)) {
  
  # Build parameter list for this site
  # Start with base parameters, then override Vmax and Km
  params_i <- base_params
  params_i$Vmax <- site_profiles$Vmax[i]
  params_i$Km   <- site_profiles$Km[i]
  
  # Attach schedules to params (the ODE function reads them from here)
  params_i$dose_schedule <- dose_schedule
  params_i$meal_schedule <- meal_schedule
  
  # Solve the ODE system
  sol <- ode(
    y      = state_init,
    times  = times,
    func   = insulin_ode,
    parms  = params_i,
    method = "lsoda"      # adaptive solver for potentially stiff system
  )
  
  # Convert to data frame and label with site name
  df <- as.data.frame(sol)
  df$site <- site_profiles$site[i]
  
  all_results[[i]] <- df
}

# Combine all site results into one long-format data frame
results <- bind_rows(all_results)

# Set site as an ordered factor (for consistent plot ordering)
results$site <- factor(results$site, levels = site_profiles$site)

# Preview the structure
cat("Simulation complete:", nrow(results), "rows across", 
    length(unique(results$site)), "sites\n")
cat("Columns:", paste(names(results), collapse = ", "), "\n")
```

---

# 7. Compute PK Metrics

```{r pk_metrics}
# ============================================================================
# PHARMACOKINETIC METRICS
# ============================================================================
# For each injection site, we compute the three primary PK metrics from the
# proposal, plus glucose time-in-range:
#
#   Cmax: Peak plasma insulin concentration (Units/L)
#         The highest value P(t) reaches. Higher Cmax means a stronger
#         but more transient insulin effect.
#
#   Tmax: Time to peak concentration (hours)
#         How long after the first injection until P reaches Cmax.
#         Faster-absorbing sites (abdomen) have shorter Tmax.
#
#   AUC:  Area under the plasma insulin curve (Units·hr/L)
#         Total systemic insulin exposure over 24 hours. Computed via
#         the trapezoidal rule. Higher AUC = more total insulin action.
#
#   TIR:  Time in glucose range 70–180 mg/dL (%)
#         Clinical metric — what fraction of the day is glucose in the
#         acceptable range? Higher = better glycemic control.
# ============================================================================

pk_table <- results %>%
  group_by(site) %>%
  summarise(
    # Peak plasma concentration
    Cmax = max(P, na.rm = TRUE),
    
    # Time of peak (first occurrence)
    Tmax = time[which.max(P)],
    
    # AUC via trapezoidal rule:
    # sum of (average of consecutive P values) * (time step)
    AUC = sum(diff(time) * (P[-1] + P[-n()]) / 2),
    
    # Glucose time-in-range (70–180 mg/dL)
    TIR = mean(G >= 70 & G <= 180) * 100,
    
    # Additional: peak glucose excursion
    G_max = max(G, na.rm = TRUE),
    G_min = min(G, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

pk_table %>%
  kable(
    caption = "Pharmacokinetic Metrics by Injection Site",
    col.names = c("Site", "C_max (U/L)", "T_max (hr)", "AUC (U·hr/L)",
                  "TIR (%)", "G_max (mg/dL)", "G_min (mg/dL)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(5, color = ifelse(pk_table$TIR > 70, "green", "orange"))
```

---

# 8. Visualizations

## 8.1 Three-Axis Phase Portrait (3D Trajectory)

```{r phase_portrait_3d, fig.height=7}
# ============================================================================
# 3D PHASE PORTRAIT
# ============================================================================
# This plot shows the trajectory of the system through its three-dimensional
# state space (S, P, G). Each point is the system state at a moment in time.
#
# Why this matters:
# - The shape of the trajectory reveals the system's dynamical structure
# - Loops/spirals indicate oscillatory behavior (dose-meal cycles)
# - The "tightness" of the trajectory shows how quickly the system returns
#   to its attractor after perturbation (dose or meal)
# - Different sites produce differently shaped trajectories because
#   faster absorption (abdomen) creates sharper excursions in P
#
# We use plotly for interactive 3D rotation.
# ============================================================================

# Select one site for the primary 3D view (can rotate interactively)
site_3d <- "Abdomen"
df_3d <- results %>% filter(site == site_3d)

# Color by time for temporal progression
plot_ly(
  df_3d,
  x = ~S, y = ~P, z = ~G,
  type = "scatter3d",
  mode = "lines",
  line = list(
    color = ~time,
    colorscale = list(c(0, "#48bb78"), c(0.5, "#4299e1"), c(1, "#e53e3e")),
    width = 3
  ),
  hoverinfo = "text",
  text = ~paste0("t=", round(time, 2), "hr<br>S=", round(S, 2),
                 "<br>P=", round(P, 2), "<br>G=", round(G, 1))
) %>%
  layout(
    title = list(
      text = paste("3D Phase Portrait —", site_3d),
      font = list(size = 16)
    ),
    scene = list(
      xaxis = list(title = "S — SC Depot (Units)"),
      yaxis = list(title = "P — Plasma Insulin (U/L)"),
      zaxis = list(title = "G — Blood Glucose (mg/dL)"),
      camera = list(eye = list(x = 1.5, y = 1.5, z = 1.0))
    ),
    margin = list(t = 50)
  )
```

```{r phase_portrait_all_sites, fig.height=7}
# ============================================================================
# 3D PHASE PORTRAIT — ALL SITES OVERLAID
# ============================================================================
# Overlaying all four sites on the same 3D plot lets us directly compare
# how the trajectory shape changes with injection site.
# ============================================================================

fig <- plot_ly()

for (i in 1:nrow(site_profiles)) {
  df_i <- results %>% filter(site == site_profiles$site[i])
  fig <- fig %>%
    add_trace(
      data = df_i,
      x = ~S, y = ~P, z = ~G,
      type = "scatter3d",
      mode = "lines",
      line = list(color = site_profiles$color[i], width = 3),
      name = site_profiles$site[i]
    )
}

fig %>%
  layout(
    title = list(text = "3D Phase Portrait — All Injection Sites", font = list(size = 16)),
    scene = list(
      xaxis = list(title = "S (SC Depot)"),
      yaxis = list(title = "P (Plasma Insulin)"),
      zaxis = list(title = "G (Blood Glucose)"),
      camera = list(eye = list(x = 1.6, y = 1.4, z = 0.9))
    )
  )
```


## 8.2 Time Series — All State Variables by Site

```{r time_series, fig.height=10, fig.width=11}
# ============================================================================
# TIME SERIES: S(t), P(t), G(t) FOR ALL INJECTION SITES
# ============================================================================
# Three-panel time series showing how each state variable evolves over 24 hours
# under the same dosing/meal schedule but different injection sites.
#
# Key things to observe:
#   Panel 1 (S): Slower sites (thigh) show higher depot peaks and slower
#     drainage — insulin accumulates more between doses.
#   Panel 2 (P): Faster sites (abdomen) produce higher, sharper plasma peaks.
#     Slower sites produce flatter, more prolonged profiles.
#   Panel 3 (G): The glucose response integrates both insulin kinetics and
#     meal timing. Look for how post-meal glucose excursions differ by site.
# ============================================================================

# Define site colors for consistent use across all plots
site_colors <- setNames(site_profiles$color, site_profiles$site)

# --- Panel 1: Subcutaneous Depot S(t) ---
p_S <- ggplot(results, aes(x = time, y = S, color = site)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  scale_color_manual(values = site_colors) +
  labs(
    title = "Subcutaneous Depot  S(t)",
    subtitle = "Insulin remaining at injection site — slower sites retain more",
    x = NULL, y = "S (Units)", color = "Site"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )

# --- Panel 2: Plasma Insulin P(t) ---
p_P <- ggplot(results, aes(x = time, y = P, color = site)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  scale_color_manual(values = site_colors) +
  labs(
    title = "Plasma Insulin  P(t)",
    subtitle = "Active circulating insulin — faster sites produce sharper peaks",
    x = NULL, y = "P (Units/L)", color = "Site"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )

# --- Panel 3: Blood Glucose G(t) ---
# Add reference lines for the clinically acceptable range (70–180 mg/dL)
p_G <- ggplot(results, aes(x = time, y = G, color = site)) +
  geom_hline(yintercept = c(70, 180), linetype = "dashed", color = "grey60", linewidth = 0.5) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 70, ymax = 180,
           fill = "green", alpha = 0.04) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  scale_color_manual(values = site_colors) +
  annotate("text", x = 23.5, y = 175, label = "180", color = "grey50", size = 3, hjust = 1) +
  annotate("text", x = 23.5, y = 75, label = "70", color = "grey50", size = 3, hjust = 1) +
  labs(
    title = "Blood Glucose  G(t)",
    subtitle = "Glycemic response — green band = target range (70–180 mg/dL)",
    x = "Time (hours)", y = "G (mg/dL)", color = "Site"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )

# Compose into a 3-panel figure using patchwork
p_S / p_P / p_G +
  plot_annotation(
    title = "24-Hour Time Series by Injection Site",
    subtitle = "Same dosing schedule, same meals — only absorption parameters vary",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12)
    )
  )
```


## 8.3 Absorption Rate Dynamics

```{r absorption_rate, fig.height=5}
# ============================================================================
# ABSORPTION RATE: k_abs(S) OVER TIME
# ============================================================================
# This shows the instantaneous absorption flux from the SC depot to plasma.
# The Michaelis-Menten nonlinearity is visible here:
#   - Immediately after injection: S is large → kabs saturates near Vmax
#   - As the depot drains: kabs transitions to first-order (linear in S)
#   - Between doses: kabs drops to near zero (depot nearly empty)
#
# Faster sites (abdomen, high Vmax) show taller but narrower absorption
# peaks. Slower sites (thigh, low Vmax) show shorter but wider peaks.
# The AREA under each peak is similar (same dose), but the SHAPE differs.
# ============================================================================

ggplot(results, aes(x = time, y = kabs, color = site, fill = site)) +
  geom_area(alpha = 0.15, position = "identity") +
  geom_line(linewidth = 0.7) +
  scale_color_manual(values = site_colors) +
  scale_fill_manual(values = site_colors) +
  labs(
    title = "Absorption Rate  k_abs(S) Over Time",
    subtitle = "Michaelis-Menten flux from depot → plasma (saturates at Vmax post-injection)",
    x = "Time (hours)", y = "Absorption Rate (U/hr)",
    color = "Site", fill = "Site"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    legend.position = "bottom"
  )
```


## 8.4 Endogenous Secretion & Feedback Dynamics

```{r feedback_dynamics, fig.height=7}
# ============================================================================
# ENDOGENOUS SECRETION E(G) AND INHIBITORY FACTOR C(t)
# ============================================================================
# These two panels reveal the internal feedback loop:
#
# Top panel: E(G) — how β-cell secretion responds to glucose changes.
#   When glucose spikes after a meal, E(G) rises. But this is modulated
#   by C(t), which suppresses secretion when plasma insulin is already high.
#
# Bottom panel: C(t) — the inhibitory factor.
#   After injection, plasma insulin P rises → C(t) drops (suppression).
#   Between doses, P falls → C(t) recovers toward 1 (full secretion).
#
# The interplay between these two functions is what prevents the P ↔ G
# system from oscillating out of control.
# ============================================================================

# Select one site for clarity
df_feedback <- results %>% filter(site == "Abdomen")

p_EG <- ggplot(df_feedback, aes(x = time)) +
  geom_line(aes(y = E_G), color = "#805ad5", linewidth = 0.9) +
  geom_line(aes(y = U_PG), color = "#38b2ac", linewidth = 0.7, linetype = "dashed") +
  labs(
    title = "Endogenous Secretion E(G)  &  Utilization U(P,G)",
    subtitle = "Purple = β-cell secretion rate | Teal dashed = glucose uptake rate",
    x = NULL, y = "Rate"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )

p_Ct <- ggplot(df_feedback, aes(x = time, y = C_t)) +
  geom_line(color = "#e53e3e", linewidth = 0.9) +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey50") +
  annotate("text", x = 23, y = 0.53, label = "50% suppression", 
           color = "grey50", size = 3, hjust = 1) +
  ylim(0, 1) +
  labs(
    title = "Inhibitory Factor  C(t)",
    subtitle = "Drops when plasma insulin is high → suppresses endogenous secretion",
    x = "Time (hours)", y = "C(t)  [0 = fully suppressed, 1 = no suppression]"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )

p_EG / p_Ct +
  plot_annotation(
    title = "Feedback Loop Dynamics (Abdomen Site)",
    theme = theme(plot.title = element_text(face = "bold", size = 15))
  )
```


## 8.5 Phase Planes (2D Projections)

```{r phase_planes, fig.height=5, fig.width=12}
# ============================================================================
# 2D PHASE PLANE PROJECTIONS
# ============================================================================
# While the 3D portrait shows the full trajectory, 2D projections are
# easier to read and reveal specific pairwise relationships:
#
#   S vs P: Shows the absorption-clearance relationship.
#           Dose events push the system to high S, then it drains toward
#           high P as absorption converts depot to plasma insulin.
#
#   P vs G: Shows the insulin-glucose control loop.
#           Post-meal: G rises → P rises (via secretion + injection).
#           Post-peak: P drives G down via utilization.
#           The shape of these loops differs by site.
#
#   S vs G: Shows how depot status relates to glycemic state.
#           Not directly coupled, but correlated through P.
# ============================================================================

p_SP <- ggplot(results, aes(x = S, y = P, color = site)) +
  geom_path(linewidth = 0.6, alpha = 0.7) +
  scale_color_manual(values = site_colors) +
  labs(title = "S vs P", subtitle = "Depot → Plasma absorption path",
       x = "S (Depot)", y = "P (Plasma)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

p_PG <- ggplot(results, aes(x = P, y = G, color = site)) +
  geom_path(linewidth = 0.6, alpha = 0.7) +
  geom_hline(yintercept = c(70, 180), linetype = "dashed", color = "grey70", linewidth = 0.4) +
  scale_color_manual(values = site_colors) +
  labs(title = "P vs G", subtitle = "Insulin-glucose control loop",
       x = "P (Plasma Insulin)", y = "G (Glucose)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

p_SG <- ggplot(results, aes(x = S, y = G, color = site)) +
  geom_path(linewidth = 0.6, alpha = 0.7) +
  geom_hline(yintercept = c(70, 180), linetype = "dashed", color = "grey70", linewidth = 0.4) +
  scale_color_manual(values = site_colors) +
  labs(title = "S vs G", subtitle = "Depot status vs glycemic state",
       x = "S (Depot)", y = "G (Glucose)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))

p_SP + p_PG + p_SG +
  plot_annotation(
    title = "2D Phase Plane Projections — All Sites",
    theme = theme(plot.title = element_text(face = "bold", size = 15))
  )
```


## 8.6 PK Metrics Comparison

```{r pk_comparison, fig.height=5, fig.width=11}
# ============================================================================
# PK METRICS BAR CHARTS
# ============================================================================
# Visual comparison of the key pharmacokinetic metrics across sites.
# These are the primary outputs of the model — the metrics from the proposal
# (Cmax, Tmax, AUC) plus the clinical glucose time-in-range.
# ============================================================================

# Reshape for faceted plotting
pk_long <- pk_table %>%
  select(site, Cmax, Tmax, AUC, TIR) %>%
  pivot_longer(-site, names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric,
    levels = c("Cmax", "Tmax", "AUC", "TIR"),
    labels = c("C_max (U/L)", "T_max (hr)", "AUC (U·hr/L)", "Time in Range (%)")
  ))

ggplot(pk_long, aes(x = site, y = value, fill = site)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = round(value, 1)), vjust = -0.4, size = 3.2, fontface = "bold") +
  facet_wrap(~ metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = site_colors) +
  labs(
    title = "Pharmacokinetic Metrics by Injection Site",
    subtitle = "Same dose, same meals — site affects absorption kinetics only",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 35, hjust = 1, size = 9),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey50", size = 11)
  )
```


## 8.7 Sensitivity Analysis: Dose Size

```{r sensitivity_dose, fig.height=6, fig.width=10}
# ============================================================================
# SENSITIVITY ANALYSIS: VARYING DOSE SIZE
# ============================================================================
# How does changing the dose magnitude affect the PK profile?
# We run the abdomen site at 5, 10, 15, and 20 Units per injection
# (all other parameters constant) and overlay the plasma insulin and
# glucose time series.
#
# This shows:
#   - How Cmax scales with dose (nonlinearly due to MM absorption)
#   - Whether higher doses improve glucose control or cause hypoglycemia
#   - The therapeutic window between efficacy and over-dosing
# ============================================================================

dose_levels <- c(5, 10, 15, 20)
dose_results <- list()

for (d in dose_levels) {
  # Scale all doses proportionally
  ds <- dose_schedule
  ds$amount <- ds$amount * (d / 10)  # 10 is the baseline
  
  params_d <- base_params
  params_d$Vmax <- 18   # Abdomen
  params_d$Km   <- 30
  params_d$dose_schedule <- ds
  params_d$meal_schedule <- meal_schedule
  
  sol <- ode(y = state_init, times = times, func = insulin_ode,
             parms = params_d, method = "lsoda")
  df <- as.data.frame(sol)
  df$dose_label <- paste0(d, " U")
  dose_results[[length(dose_results) + 1]] <- df
}

dose_df <- bind_rows(dose_results)
dose_df$dose_label <- factor(dose_df$dose_label, 
                              levels = paste0(dose_levels, " U"))

p_dose_P <- ggplot(dose_df, aes(x = time, y = P, color = dose_label)) +
  geom_line(linewidth = 0.8) +
  scale_color_brewer(palette = "YlOrRd") +
  labs(title = "Plasma Insulin — Dose Sensitivity",
       subtitle = "Abdomen site, varying bolus size",
       x = NULL, y = "P (U/L)", color = "Dose") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none")

p_dose_G <- ggplot(dose_df, aes(x = time, y = G, color = dose_label)) +
  geom_hline(yintercept = c(70, 180), linetype = "dashed", color = "grey60") +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 70, ymax = 180,
           fill = "green", alpha = 0.04) +
  geom_line(linewidth = 0.8) +
  scale_color_brewer(palette = "YlOrRd") +
  labs(title = "Blood Glucose — Dose Sensitivity",
       subtitle = "Higher doses → more aggressive glucose lowering (risk of hypo)",
       x = "Time (hours)", y = "G (mg/dL)", color = "Dose") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")

p_dose_P / p_dose_G
```

---

# 9. Summary

```{r summary_text, results='asis'}
# ============================================================================
# SUMMARY
# ============================================================================
cat("
## Key Findings from the Preliminary Model

**1. Site-dependent absorption produces measurably different PK profiles.**
The abdomen (highest Vmax, lowest Km) produces the tallest, sharpest plasma 
insulin peaks, while the thigh produces broader, flatter profiles with the 
same total dose.

**2. Faster absorption improves postprandial glucose control.**
The abdomen site achieves lower peak glucose excursions and higher time-in-range
because insulin arrival is better aligned with meal glucose appearance.

**3. Slower sites accumulate more insulin in the depot.**
The thigh shows visible depot buildup between doses — insulin from one injection 
has not fully cleared before the next arrives. This could contribute to 
unpredictable delayed absorption.

**4. The feedback loop (E(G) · C(t)) provides endogenous damping.**
When exogenous insulin raises P, the inhibitory factor C(t) suppresses 
endogenous secretion, preventing insulin stacking. This feedback is visible 
in the C(t) time series as dips following each injection.

**5. Dose sensitivity is nonlinear.**
Due to Michaelis-Menten absorption, doubling the dose does not double Cmax.
The depot saturates at Vmax, so excess insulin is released more slowly —
a built-in buffering mechanism.

### Next Steps (per phased plan)
- **Phase 2**: Validate glucose compartment coupling against published 
  postprandial glucose profiles.
- **Phase 3**: Systematically sweep injection-meal timing offsets and 
  dose-splitting strategies across all four sites.
")
```

---

*Model equations and parameter values are preliminary and drawn from published 
literature (Gradel et al. 2018, Koivisto & Felig 1980). Patient-specific 
calibration and formal validation are planned for Phase 3.*
